### **Hypothesis Testing:**
\label{sec:analysis}
```{r load library, message=FALSE}
library(tidyverse)
library(knitr)
```

Hypothesis test for independence of a difference in medians using permutation is
a suitable test to answer our main question, given that the suitable estimator 
is median and each data point is collected independently from several testing 
stations and independent time unit. Our pre-specified significance level is 
$\alpha$ = 0.05 (the probability for us to reject the null hypothesis)

-   Null Hypothesis($H_0$): The measurement of PM2.5 in Beijing from time_B does 
not show any sign of improvement comparing to time_A. (Point estimate of PM2.5 
in time_B  $\geq$ the point estimate of PM2.5 in time_A)

-   Alternative Hypothesis($H_a$): The measurement of PM2.5 in Beijing from 
time_B shows an improvement comparing to time_A. (Point estimate of PM2.5 in 
time_B is $<$  the point estimate of PM2.5 in time_A)

```{r load p-value, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
pvalue <- read_csv(here("results", "pvalue.csv"))
```
After conducting the permutation test in the difference of medians, we get a 
p-value as:
```{r output p-value}
pvalue[2][[1]]
```
which is greater than the significance level $\alpha$ = 0.05. 

The result indicates that we do not have enough statistical evidence to reject 
the null hypothesis, $H_0$. Hence, there is no statistically significant 
difference between the median PM2.5 measurement for time_B and time_A in Beijing.


