---
title: "Exploratory data analysis of Beijing air quality data set"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE)
library(here)
library(tidyverse)
library(knitr)
library(ggplot2)
library(ggthemes)
library(scales)
library(cowplot)
theme_set(theme_minimal())
set.seed(2020)
```

# Summary of the data set

## Data Import

Data are downloaded and unzipped by [python script](`r here("src", "download_data.py")`). 12 csv files are used in this project. All in the same format with one file representing one Beijing district. All data are combined and stored in`air_data`.

```{r load data}
csv_list <- dir(here("data", "raw","PRSA_Data_20130301-20170228"))

air_data <- data.frame()
for (file_name in csv_list) {
        air_data <-
                rbind(air_data, 
                      read_csv(here("data", 
                                    "raw", 
                                    "PRSA_Data_20130301-20170228", 
                                    file_name), 
                               show_col_types = FALSE))
}

head(air_data)
```

## Study the data

The data set used in this project is an hourly air pollutants data from 12 nationally-controlled air-quality monitoring sites in Beijing, China from 1 March 2013 to 28 February 2017. The air-quality data is from the Beijing Municipal Environmental Monitoring Center. It was sourced from the UCI Machine Learning Repository [here](https://archive-beta.ics.uci.edu/ml/datasets/beijing+multi+site+air+quality+data).

Below we show the structure of each features in the data set.

```{r study data}
str(air_data, give.attr = FALSE)
```

Each row in the data set represents a measurements of air pollutants (e.g., PM2.5, PM10, CO) at specific date and time in 12 different district in Beijing, including Aotizhongxin, Changping, Dingling and so on. The meteorological data in each air-quality site are matched with the nearest weather station from the China Meteorological Administration. There are `420,768` observations in the data set, and `18` features. Below we show the description and missing value of each features in the data set.

```{r missing value}
missing_values <- data.frame(apply(is.na(air_data), 2, sum))

missing_values <- cbind(
        missing_values,
        c(
                'row number',
                'year of data in this row',
                'month of data in this row',
                'day of data in this row',
                'hour of data in this row',
                'PM2.5 concentration (ug/m^3)',
                'PM10 concentration (ug/m^3)',
                'SO2 concentration (ug/m^3)',
                'NO2 concentration (ug/m^3)',
                'CO concentration (ug/m^3)',
                'O3 concentration (ug/m^3)',
                'temperature (degree Celsius)',
                'pressure (hPa)',
                'dew point temperature (degree Celsius)',
                'precipitation (mm)',
                'wind direction',
                'wind speed (m/s)',
                'name of the air-quality monitoring site'
        )
)

colnames(missing_values) <- c("# of missing values", "description")

missing_values <-
        missing_values |> select("description", "# of missing values")

knitr::kable(missing_values, "pipe")
```

(Histogram + short description)

# Data Wrangling

We are interested to determine how PM2.5 has changed over two time periods - between **March 1 2013 to Feb 28 2015**, and **March 1 2015 to Feb 28 2017**. As such, we will drop all irrelevant columns, and only kept year, month, PM2.5. We have also created a derived column 'class', which will act as our target variable.

```{r spilt data}
air_data_processed <- air_data |>
  select(year, month, PM2.5) |>
  mutate(class = ifelse((year >=2013 & year <=2014) | (year == 2015 & month <=2),
                        'time_A',
                        'time_B'))

air_data_processed
```

```{r data summary year}
summary_by <- function(col) {
        air_data_summary <- air_data_processed |>
                group_by( {{ col }} ) |>
                summarise(mean := round(mean(PM2.5, na.rm = TRUE)))
        
        air_data_summary_transpose <-
                as.data.frame(t(as.matrix(air_data_summary$mean)), row.names = 'mean')
        
        colnames(air_data_summary_transpose) <-
                unlist(air_data_summary[1])
        
        air_data_summary_transpose
}

knitr::kable(summary_by(year), "pipe")
```

Figure 1. Table of Sample mean of PM2.5 by year

```{r data summary month}
knitr::kable(summary_by(month), "pipe")
```

Figure 2. Table of Sample mean of PM2.5 by month

# Exploratory analysis on the training data set

We will perform exploratory data analysis steps to highlight the distribution of PM2.5 data points for both sample A and sample B.

## Heading

(hist?)

```{r EDA hisogram plot}
sample_A_B_boxplot <- air_data_processed |>
  group_by(class) |>
  ggplot(aes(x=PM2.5, y=class, fill = class)) +
  geom_boxplot(names=c('Sample A', 'Sample B')) +
  geom_point(stat = 'summary', fun = 'mean', color = 'black') +
  labs(title = 'Sample A has a higher median and mean value than Sample B',
       subtitle = 'Both Sample A and Sample B are heavily right skewed') +
  theme(axis.title.y=element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.x=element_blank(),
        axis.text.x = element_blank())

sample_A_B_distribution <- air_data_processed |>
  group_by(class) |>
  ggplot(aes(x=PM2.5, fill = class)) +
  geom_density(alpha = 0.4) +
  theme(axis.title.y=element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())

plot_grid(sample_A_B_boxplot, sample_A_B_distribution, ncol=1)

```

Figure 2. Boxplot and Density plot for both samples' PM2.5 distribution. (Black dots represent the mean PM2.5 value of each sample).

Both samples are heavily right skewed. Sample A has a higher median and mean PM2.5 value than Sample B, and its data distribution is wider than B. It is fair to conclude that the median and mean values are too similar to make a definitive statement as to whether the air quality has changed over these two time frames.

## Heading

```{r EDA density plot}

sample_A_B_distribution <- air_data_processed |>
  group_by(class) |>
  ggplot(aes(x=log(PM2.5), fill = class)) +
  geom_density(alpha = 0.4) +
  theme(axis.title.y=element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
 sample_A_B_distribution
```

## Heading 3

Confident 95% + ticket
