---
title: "Exploratory data analysis of Beijing air quality data set"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE)
library(here)
library(tidyverse)
library(knitr)
library(ggplot2)
library(ggthemes)
library(scales)
library(cowplot)
theme_set(theme_minimal())
set.seed(2020)
```

# Summary of the data set

## Data Import

```{r load_data}
csv_list <- dir(here("data", "raw","PRSA_Data_20130301-20170228"))

air_data <- data.frame()
for (file_name in csv_list) {
        air_data <-
                rbind(air_data, 
                      read_csv(here("data", 
                                    "raw", 
                                    "PRSA_Data_20130301-20170228", 
                                    file_name), 
                               show_col_types = FALSE))
}

head(air_data)
```

The data set used in this project is an hourly air pollutants data from 12 nationally-controlled air-quality monitoring sites in Beijing, China from 1 March 2013 to 28 February 2017. The air-quality data is from the Beijing Municipal Environmental Monitoring Center. It was sourced from the UCI Machine Learning Repository [here](https://archive-beta.ics.uci.edu/ml/datasets/beijing+multi+site+air+quality+data).

Each row in the data set represents a measurements of air pollutants (e.g., PM2.5, PM10, CO) at specific date and time in 12 different district in Beijing, including Aotizhongxin, Changping, Dingling and so on. The meteorological data in each air-quality site are matched with the nearest weather station from the China Meteorological Administration. There are `420,768` observations in the data set, and `18` features. Below we show the structure of each features in the data set.

## Study the data

```{r study_data}
str(air_data)
```

# Data Wrangling 

We are interested to determine how PM2.5 has changed over two time periods - between March 1 2013 to Feb 28 2015, and March 1 2015 to Feb 28 2017. As such, we will drop all irrelevant columns, and only keep year, month, PM2.5. We have also created a derived column 'class', which will act as our target variable.

```{r data wrangling}
air_data_processed <- air_data |>
  select(year, month, PM2.5)
```

```{r Creating new column called class and re-binding it}
# Isolate Sample A
sample_A_2015 <- air_data_processed |>
  filter(year == 2015,
         month <= 2)

Sample_A_2013_2014 <- air_data_processed |>
  filter(year >= 2013 & year <= 2014)

Sample_A <- rbind(Sample_A_2013_2014,sample_A_2015) |>
  mutate(class = 'Sample_A')

# Isolate Sample B
sample_B_2015 <- air_data_processed |>
  filter(year == 2015,
         month > 2)

Sample_B_2016_2017 <- air_data_processed |>
  filter(year>2015)

Sample_B <- rbind(sample_B_2015, Sample_B_2016_2017) |>
  mutate(class = 'Sample_B')

# Processed Dataframe
air_data_processed <- rbind(Sample_A, Sample_B)
air_data_processed
```

Figure 1. Processed dataset with new derived column (sample).

# Exploratory analysis on the training data set

We will perform exploratory data analysis steps to highlight the distribution of PM2.5 data points for both sample A and sample B.

```{r EDA boxplot and density plot}
sample_A_B_boxplot <- air_data_processed |>
  group_by(class) |>
  ggplot(aes(x=PM2.5, y=class, fill = class)) +
  geom_boxplot(names=c('Sample A', 'Sample B')) +
  geom_point(stat = 'summary', fun = 'mean', color = 'black') +
  labs(title = 'Sample A has a higher median and mean value than Sample B',
       subtitle = 'Both Sample A and Sample B are heavily right skewed') +
  theme(axis.title.y=element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.x=element_blank(),
        axis.text.x = element_blank())

sample_A_B_distribution <- air_data_processed |>
  group_by(class) |>
  ggplot(aes(x=PM2.5, fill = class)) +
  geom_density(alpha = 0.4) +
  theme(axis.title.y=element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())

plot_grid(sample_A_B_boxplot, sample_A_B_distribution, ncol=1)

```

Figure 2. Boxplot and Density plot for both samples' PM2.5 distribution. (Black dot represents the mean PM2.5 value of each sample).

Both samples are heavily right skewed. Sample A has a higher median and mean PM2.5 value than Sample B, and its data distribution is wider than B. It is fair to conclude that the median and mean values are too similar to make a definitive statement as to whether the air quality has changed over these two time frames.
