---
title: "Exploratory data analysis of Beijing air quality data set"
author: "Jacqueline Chong, Junrong Zhu, Macy Chan, Vadim Taskaev"
output: 
  github_document:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE)
library(here)
library(tidyverse)
library(knitr)
library(ggplot2)
library(ggthemes)
library(scales)
library(cowplot)
theme_set(theme_minimal())
set.seed(2020)
```

## Data Import

The data is downloaded and unzipped by a [python script](https://github.com/UBC-MDS/DSCI_522_Beijing_Air_Quality/blob/main/src/download_data.py). 12 csv files were used in this project. Each file represents one of 12 nationally-controlled air-quality monitoring sites, which captured information **between March 2013 and February 2017**. They are all in the same format (.csv files), and are merged and stored in `air_data`.

```{r load data}
csv_list <- dir(here("data", "raw","PRSA_Data_20130301-20170228"))

air_data <- data.frame()
for (file_name in csv_list) {
  air_data <- rbind(air_data,
                    read_csv(
                      here("data",
                           "raw", 
                           "PRSA_Data_20130301-20170228", 
                           file_name), 
                      show_col_types = FALSE))
}

head(air_data) |>
  kable(caption = "Table 1. Preview of original datset")
```

## Study the data

This data set contains hourly air pollutants data from 12 nationally-controlled air-quality monitoring sites in Beijing, China from **1 March 2013 to 28 February 2017**. The air-quality data is retrieved from the Beijing Municipal Environmental Monitoring Center. It was sourced from the UCI Machine Learning Repository [here](https://archive-beta.ics.uci.edu/ml/datasets/beijing+multi+site+air+quality+data).

This is the structure of each features in the data set.

```{r study data}
str(air_data, give.attr = FALSE)
```

Table 2: Structure of our data set `air_data`

Each row in the data set represents a measurement of air pollutants (e.g., PM2.5, PM10, CO) at specific year, month, and day in 12 different district in Beijing, including Aotizhongxin, Changping, Dingling. The meteorological data in each air-quality site are matched with the nearest weather station from the China Meteorological Administration. There are `420,768` observations and `18` features in the data set.

Below we show the description and missing value of each features in the data set.

```{r missing value}
missing_values <- data.frame(apply(is.na(air_data), 2, sum))

missing_values <- cbind(
        missing_values,
        c(
                'Row number',
                'Year of data in this row',
                'Month of data in this row',
                'Day of data in this row',
                'Hour of data in this row',
                'PM2.5 concentration (ug/m^3)',
                'PM10 concentration (ug/m^3)',
                'SO2 concentration (ug/m^3)',
                'NO2 concentration (ug/m^3)',
                'CO concentration (ug/m^3)',
                'O3 concentration (ug/m^3)',
                'Temperature (degree Celsius)',
                'Pressure (hPa)',
                'Dew point temperature (degree Celsius)',
                'Precipitation (mm)',
                'Wind direction',
                'Wind speed (m/s)',
                'Name of the air-quality monitoring site'
        )
)

colnames(missing_values) <- c("# of missing values", "Description")

missing_values <-
        missing_values |> select("Description", "# of missing values")

missing_values |>
  kable(caption = "Table 3. Description and instances of missing values in the original dataset")
```

As we are interested to determine how PM2.5 levels have changed over two time periods - between **March 1 2013 to Feb 28 2015**, and **March 1 2015 to Feb 28 2017**, we would first like to see if there are any micro changes associated with how these levels shift across years and months. Here we are going to create two summary tables on yearly and monthly PM2.5 values.

```{r data summary year}
summary_by <- function(col) {
  air_data_summary <- air_data |>
    
    group_by( {{ col }} ) |>
    summarise(mean := round(mean(PM2.5, na.rm = TRUE)),
              min := round(min(PM2.5, na.rm = TRUE)),
              max := round(max(PM2.5, na.rm = TRUE)))
        
    air_data_summary_transpose <- as.data.frame(
      t(as.matrix(air_data_summary)), 
      row.names = c('year','mean','min', 'max'))
    
    names(air_data_summary_transpose) <- lapply(
      air_data_summary_transpose[1, ], as.character)
    
    air_data_summary_transpose <- air_data_summary_transpose[-1,] 

    air_data_summary_transpose
}

summary_by(year) |>
  kable(caption = "Table 4. Mean, max and min of PM2.5 per year")
```

Looking at the statistic values on **Table 4**, 2016 had the largest range of PM2.5 levels (from 72(ug/m^3) to 999(ug/m^3)), while 2015 had the lowest `pm_max` value but its mean was the same as in 2013.

```{r year figure, fig.width=10, fig.height=5}
year_figure<- air_data |> 
              select(PM2.5, year) |> 
              drop_na(PM2.5) |> 
              group_by(year) |> 
              summarize(pm_mean = mean(PM2.5),
                        pm_max = max(PM2.5),
                        pm_min = min(PM2.5)) |> 
              pivot_longer(cols = pm_mean:pm_min,
                           names_to = "summary_values") |> 
              ggplot(aes(x = year, y = value, color = summary_values))+
              geom_line()+
              labs(x = "Year", y = "PM2.5 measurements", color = "Summary Values", 
                   title = "Yearly Summary Line Plot")

year_figure
```

Figure 1. Yearly Summary Line Plot of PM2.5 Measurements, illustrating the evolution of PM2.5 mean PM2.5 measurements (in green) as well as minimum and maximum PM2.5 measurements (in blue and red) between 2013 and 2017 (inclusive). This figure displays the trend a relatively steady mean PM2.5 measurements, despite a small decrease in 2016 and subsequent substantial increase in 2017, and points to relatively large variability of maximum measurement observations.

As seen from **Figure 1** above, the variation in the **max** PM2.5 measurements across the years is the most obvious comparing to the changes in the overall `pm_min` and `pm_mean` values.

```{r data summary month}
summary_by(month)|>
  kable(caption = "Table 5. Mean, max and min of PM2.5 per month")
```

```{r month figure, fig.width=10, fig.height=5}
month_figure <- air_data |> 
                select(PM2.5, month) |> 
                drop_na(PM2.5) |> 
                group_by(month) |> 
                summarize(pm_mean = mean(PM2.5),
                          pm_max = max(PM2.5),
                          pm_min = min(PM2.5)) |> 
                pivot_longer(cols = pm_mean:pm_min,
                             names_to = "summary_values") |>
                ggplot(aes(x = month, 
                           y = value, 
                           color = summary_values))+
                geom_line()+
                labs(x = "Month", y = "PM2.5 measurements", color = "Summary Values", 
                     title = "Monthly Summary Line Plot") +
                scale_x_continuous(breaks=pretty_breaks())

month_figure
```

Figure 2. Monthly Summary Line Plot of PM2.5 Measurements. Unlike Figure 1, Figure 2 displays the aggregate monthly average of mean, minimum, and maximum PM2.5 measurements across the entire data set time series (in green, blue, and red, respectively). This figure reflects a general monthly trend whereby both mean and maximum PM2.5 measurements appears most elevated in winter months, and are at their lowest in the month of August. The observed variability of maximum PM2.5 measurements month-to-month is most noticeable throughout the months of February through June.

As seen from Figure 2, `pm_mean` is generally at the lowest in August and September and overall the variation in max surpasses the changes in mean and min values.

In general, we can not conclude significant differences among these statistical summary values to answer the main question: **Do PM2.5 measurements in Beijing, China collected from 2013 to 2017 show any sign of improvement?**. As such, we will perform further data wrangling and exploration to find out more representative characteristics of our data.

## Data Wrangling

Since we observe only slight changes in yealy PM2.5 measurements, we have chosen to focus our analysis on determining how PM2.5 measurements have changed between two time periods - between **March 1 2013 to Feb 28 2015**, and **March 1 2015 to Feb 28 2017**. Hereafter in this project, we will refer to first time period **March 1 2013 to Feb 28 2015** as `time_A` and second time period **March 1 2015 to Feb 28 2017** as `time_B` in both visualization and code scripts after we processed the raw data.

We set out to answer these questions from our proposal:

1.  How does the distribution of air quality data observations look like? Are they normally distributed?
2.  If we split the data into sub data sets based on time frames, will there be any overlap in the ranges of the samples?

Before we answer these sub-EDA questions, we dropped all irrelevant columns, and only kept **year**, **month**, **PM2.5**. Since we observed that there are `8739 missing values` in the combined raw data from **Table 3** (Description and instances of missing values), we will drop all the missing PM2.5 value rows. Furthermore, we are going to create a derived column `class`, which contains our explanatory variables `time_A` and `time_B`.

```{r spilt data}
air_data_processed <- air_data |>
  select(year, month, PM2.5) |>
  drop_na(PM2.5) |>
  mutate(class = ifelse((year >=2013 & year <=2014) | (year == 2015 & month <=2),
                        'time_A',
                        'time_B'))

air_data_processed
```

Table 6: Preview of the `air_data_processed` data frame.

## Distribution of data points for each class

```{r counts for each class}
time_A_sum <- sum(air_data_processed$class == 'time_A')
time_B_sum <- sum(air_data_processed$class == 'time_B')

table_distribution_count <- data.frame(Class = c('time_A', 'time_B'),
                                       'Sum of rows' = c(time_A_sum, time_B_sum))

table_distribution_count |>
  kable(caption = "Table 7. Sum of rows in each class.")
```

Both classes are equally distributed, even after dropping the missing PM2.5 values present in our data set. As such, we are not concern about class imbalance which could lead to statistical parity.

## Distribution of PM2.5 levels across time_A and time_B

Having verified that both classes are balanced, we move on to observe the distribution between both `time_A` and `time_B`.

```{r combined_distribution_plot, fig.width=8, fig.height=6}

time_A_B_boxplot <- air_data_processed |>
  group_by(class) |>
  ggplot(aes(x=PM2.5, y=class, fill = class)) +
  geom_boxplot(outlier.alpha = 0.1,
               outlier.fill = NULL) +
  geom_point(stat = 'summary', fun = 'mean', color = 'black') +
  labs(title = 'Box Plot and Histogram PM2.5 Distribution',
       subtitle = 'plots over both Time Frame(time_A, time_B)') +
  labs(x="PM2.5 measurement",
       y="")+
  theme(text = element_text(size = 18))+theme_bw()

histogram <- air_data_processed |>
  ggplot(aes(x=PM2.5, fill=class)) +
  geom_histogram(bins=30) +
  facet_wrap(~class, ncol = 1) +
  labs(y="Count (of hourly observations) per \nx-axis PM2.5 measurement")+
  theme(text = element_text(size = 18))+theme_bw()


plot_grid(
  plot_grid(ggplot(),
            time_A_B_boxplot,
            rel_widths = c(1,70)),
  plot_grid(histogram,
            ggplot(),
            ncol=2,
            rel_widths = c(12,1.75)),
  ncol=1)+theme_bw()
```

Figure 3. Boxplot and Histogram for both samples' PM2.5 distribution. (Black dots represent the mean PM2.5 value of each sample)

The scale in the x-axis is identical for both plots. Both samples are heavily right skewed, as seen from numerous right-sided outliers in the boxplot and the long right tail in the histograms. `time_A` has a higher median and mean PM2.5 value than `time_B`, and its data distribution is wider than B. It is fair to conclude that the **median and mean values are too similar to make a definitive statement as to whether the air quality has changed over these two time frames**.

As our distribution is heavily skewed to the right, the mean is getting drawn to the right side. As such, we will **take median as the appropriate estimator for hypothesis testing**, since it is not sensitive to extreme values since we are taking the 50th percentile of our data.

To explore further, we are plotting a log density graph to see how the PM2.5 levels are distributed.

```{r time_A_B_distribution, fig.width=10, fig.height=5}
time_A_B_distribution <- air_data_processed |>
  group_by(class) |>
  ggplot(aes(x=log(PM2.5), fill = class)) +
  geom_density(alpha = 0.4) +
  theme(axis.title.y=element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  geom_vline(xintercept = 4.1, color = 'purple') +
  geom_vline(xintercept = 1.3, color = 'orange') +
  labs(x = 'PM2.5 level (log(ug/m^3))',
       title = 'Density Plot of PM2.5 Log-Distribution per Time Frame',
       subtitle = 'time_A has more lower and upper extreme values than time_B')
time_A_B_distribution 
```

Figure 4. Density Plot for time_A and time_B

Looking at these areas:  
**(1) on the left of the orange vertical line**  
**(2)** **the right of the purple vertical line,**

it is clear that the area under the curve for `time_A` in these two sections are larger than the area under the curve of `time_B`. This indicates `time_A` has more extreme values than `time_B`. The area under the curve of `time_B` dominates between the orange line and purple vertical line (it completely covered the area of `time_A`). However, there are overlapping areas. The distinct areas of `time_A` and `time_B` in this density plot suggests that `time_A` could possibly have a higher median than `time_B`.
